%------------------------------------------------------------------------------
\chapter*{Acknowledgements}
\label{sec:ack}
%------------------------------------------------------------------------------
Even though this thesis has my name on the front a number of people were involved in shaping its final form both in person and in spirit.

First and foremost I would like to thank my late grandfather Shanti Sarup Sehgal and grandmother Kusum Sehgal. Their constant motivation and belief in my abilities has and will always inspire me to achieve more.

I am very grateful to Professor Bernhard Ketzer for giving me an opportunity to write my thesis in his group. In spite of my many shortcomings and failures he has always been patient and supportive during the entire time and has been a role model I look up to. I would also like to thank Professor Jochen Dingfelder. I have always admired your lectures and am thankful that you agreed to be my second supervisor.

A special thanks to PhD students Michael Hösgen and Martin Hoffmann for helping with the editing of this thesis. Thank you to Michael for always answering my various questions and solving my problems without which this thesis would have never been completed. A big thanks to the whole AG Ketzer group for both the academic and moral support.

I am grateful to my father Ravi Sehgal, my mother Rajni Sehgal and my brother Sambhav for all the sacrifices they have made. Without their constant support studying at Bonn would have just remained a pipe dream. Lastly, I am thankful to the wonderful set of friends I am lucky to have. Thank you Svenja ,Georgios and Amitayus for the countless Mensa lunches which helped me remain sane throughout the thesis.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../mythesis"
%%% End:
